IML Term project paper

### 2022-12-23 paper.rmd
Removed some meta comments and references to the final report.

Let's next visualize with R [@RDCT2008] some our explanatory variables via boxplots and scatterplots to get a deeper understanding of them. For the sake of simplicity, these visualizations grouped by the phenomena being measured. In order to also preliminarily look at how these variables behave in relation to new particle formation events, we'll also look at said measurements behave in relation to variable "class4". In terms of the boxplots, the boxplots are divided into groups by variable class4, and in terms of the scatterplots, the following colors represent different classes: red represents nonevent days, blue represents days with type Ia NPF events, green represents days with type Ib NPF events and black represents days with type II NPF events:

@@

```
(Other visualizations omitted from the preliminary report.)
```{r, eval=FALSE}
cols<-c()
for(i in 1:464){
  if(npf$class4[i]=="nonevent"){
    cols<-c(cols,"red")
  } else if (npf$class4[i]=="Ia"){
    cols<-c(cols,"blue")
  } else if (npf$class4[i]=="Ib"){
    cols<-c(cols,"green")
  } else {
    cols<-c(cols, "black")
  }
}
par(mfrow=c(2,4))
for(i in 2:9){
plot(1:464,npf[,i],col=cols,main = colnames(npf)[i])
}
for(i in 2:9){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 10:11){
plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 10:11){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 12:23){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 12:23){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 24:25){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 24:25){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 26:37){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 26:37){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 38:49){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 38:49){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,5))
for(i in 50:59){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 50:59){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,4))
for(i in 60:67){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 60:67){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 68:79){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 68:79){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,2))
for(i in 80:85){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 80:85){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,5))
for(i in 86:95){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 86:95){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,2))
for(i in 96:99){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 96:99){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 100:101){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 100:101){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
```

In particular we notice that with several of the variables the boxplots for the nonevent group observations differ more from the observations for days with NPF events than the boxplots for the different NPF event groups both in terms of smaller or higher values but also spanning larger intervals (although this can partially be caused by there being observations within this group). Another notable thing is that the measurements of the same variables at different heights behave very similarly, which imply correlation between them (we take a more detailed look into this in the section *Correlation between parameters*). We also notice a few interesting sets of variables, such as nitrogen monoxide (measurements NO), potential temperature gradient (measurements PTG), rain indicator signal (measurements SWS), sulphur dioxide concentration (measurements SO2) and (measurements CS), which seem to mostly have very small standard deviations. This would imply that the shifts during the day in relation to the corresponding mean observations are on average very small.

@@

The correlation between the predictors and the value to be predicted is a very natural starting point when it comes to correlation analysis. It is also possible that the predictors themselves are correlated. Many classifiers are affected by high correlation, i.e. *collinearity* between variables. In regression models, for example, it can be difficult to differentiate between the individual effects of collinear variables on the response. Let's visualize these correlations with R package [@corrplot]

@@

Notable benefits of the logistic regression model include that according to the article "On Discriminative vs. Generative classifiers: A comparison of logistic regression and naive Bayes"[@ng2001] as a discriminative model it eventually reaches a lower asymptotic variance than comparable generative models. On top of this, according to the course material [book, p. 387], compared to for example comparable support vector classifiers, on average logistic regression models fare better under situations where the classes are not easily separable by explanatory variables.

@@

## Classifier performances (numerical)

In order to reduce knitting time for the file, we will in this report instead of expressing the full R code to calculate the values, simply report the results we received. The results for the two-class models in terms of loo-cross-validated accuracy and perplexity are:
\begin{table}
\begin{tabular}{|r|ccc|}
\hline
Model type & Loo-CV accuracy & Loo-CV perplexity&Binary accuracy on test data\\
\hline
Dummy model & 0 & 2.01&\\
Logistic regression & 0.87 & 1.36&0.84\\
Naive Bayes & 0.81 & 152.85&\\
LDA & 0.89 & 1.33&\\
QDA & 0.85 & 6.34&\\
SVM & 0.88 & -&0.84\\
Lasso & 0.89 & 1.29&0.87\\
Random Forest & 0.89 & 1.35&\\
\hline
\end{tabular}
\end{table}

## Chosen classifier, pros and cons of this particular classifier for this application

@@

We made the decision of model based on preliminary results, and though classification trees would represent an equally viable solution, we calculated those results only later on, and by that point had already calculated the results for the logistic regression model. However, for the final report we might consider transitioning to the classification tree model.

@@

(Sara: muita preprocessing-probleemoja tai vastaavia, joita haluatte nostaa)

The second problem is perhaps not as grave as the first. Especially in our challenge submission, our accuracy (0.84 for two classes and 0.68 for four) and perplexity (1.47) were not all that great. This was partially due to a coding error which made us choose a model that perhaps wasn't ideal, but there was still room for improvement even with our other models. We agreed before submitting the results that we are not aiming to win anything, and the submission itself is enough. We stand by this. We still could have done better in predicting our own results, and we set to rectify this in our final report.

One of the other frustrations includes questions with our approach in relation to cross-validation. Perhaps the original choice of using loo-CV to assess the best model wasn't ideal and eventually in order to accomodate a lower amount of folds would have required us to possibly shift our entire approach from model accuracy and perplexity estimations to choice of ideal model and interpretation of said model as a result. Eventually we used a kind of messy half-measure where we decided our ideal model based on loo-CV and then for the model we used to predict the actual test data, we ran 10-fold cross validation to assess the ideal value of the tuning parameter. This was done to acknowledge the issues with the loo-CV even if, especially as we didn't enter into the contest with this model and for example the results in the table are still based on the loo-cv ideal model to compare it with the other models, and thus we didn't do much with the results we got.

Towards the 11th of December, we also found that many of our tree-based methods perform as well or better than logistic regression. If we had found this out earlier, we might have opted for a tree-based method as our final classifier. At that point we did not want to discard the work we had done so far, and opted to try to improve our logistic regression model instead.

>>>>

Let's next visualize with R [@RDCT2008] some our explanatory variables via boxplots and scatterplots to get a deeper understanding of them. For the sake of simplicity, these visualizations are grouped by the phenomena being measured. In order to also preliminarily look at how these variables behave in relation to new particle formation events, we'll also look at how said measurements behave in relation to variable "class4". In terms of the boxplots, the boxplots are divided into groups by variable class4, and in terms of the scatterplots, the following colors represent different classes: red represents nonevent days, blue represents days with type Ia NPF events, green represents days with type Ib NPF events and black represents days with type II NPF events:

@@

```{r, eval=FALSE}
cols<-c()
for(i in 1:464){
  if(npf$class4[i]=="nonevent"){
    cols<-c(cols,"red")
  } else if (npf$class4[i]=="Ia"){
    cols<-c(cols,"blue")
  } else if (npf$class4[i]=="Ib"){
    cols<-c(cols,"green")
  } else {
    cols<-c(cols, "black")
  }
}
par(mfrow=c(2,4))
for(i in 2:9){
plot(1:464,npf[,i],col=cols,main = colnames(npf)[i])
}
for(i in 2:9){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 10:11){
plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 10:11){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 12:23){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 12:23){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 24:25){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 24:25){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 26:37){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 26:37){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 38:49){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 38:49){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,5))
for(i in 50:59){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 50:59){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,4))
for(i in 60:67){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 60:67){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,4))
for(i in 68:79){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 68:79){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(3,2))
for(i in 80:85){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 80:85){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,5))
for(i in 86:95){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 86:95){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(2,2))
for(i in 96:99){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 96:99){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
par(mfrow=c(1,2))
for(i in 100:101){
  plot(1:464,npf[,i],col=cols,main=colnames(npf)[i])
}
for(i in 100:101){
  boxplot(npf[,i]~npf$class4,main = colnames(npf)[i],names=c("Ia","Ib","II","N"))
}
```

In particular we notice that with several of the variables the boxplots for the nonevent group observations differ more from the observations for days with NPF events than the boxplots for the different NPF event groups both in terms of smaller or higher values but also spanning larger intervals (although this can partially be caused by there being observations within this group). Another notable thing is that the measurements of the same variables at different heights behave very similarly, which imply correlation between them (we take a more detailed look into this in the section *Correlation between parameters*). While looking at similar visualizations of other variables, we noticed a few interesting sets of variables such as nitrogen monoxide (measurements NO), potential temperature gradient (measurements PTG), rain indicator signal (measurements SWS), sulphur dioxide concentration (measurements SO2) and (measurements CS), which seem to mostly have very small standard deviations. This would imply that the shifts during the day in relation to the corresponding mean observations are on average very small.

@@

The correlation between the predictors and the value to be predicted is a very natural starting point when it comes to correlation analysis. It is also possible that the predictors themselves are correlated. Many classifiers are affected by high correlation, i.e. *collinearity* between variables. In regression models, for example, it can be difficult to differentiate between the individual effects of collinear variables on the response. Let's visualize these correlations with R package [@corrplot2021]

@@

Notable benefits of the logistic regression model include that according to the article "On Discriminative vs. Generative classifiers: A comparison of logistic regression and naive Bayes"[@Ng2001] as a discriminative model it eventually reaches a lower asymptotic variance than comparable generative models. On top of this, according to the course material [book, p. 387], compared to for example comparable support vector classifiers, on average logistic regression models fare better under situations where the classes are not easily separable by explanatory variables.

@@

## Classifier performances

The results for the two-class models in terms of loo-cross-validated accuracy and perplexity are:
\begin{table}
\begin{tabular}{|r|ccc|}
\hline
Model type & Loo-CV accuracy & Loo-CV perplexity&Binary accuracy on test data\\
\hline
Dummy model & 0 & 2.01&\\
Logistic regression & 0.87 & 1.36&0.84\\
Naive Bayes & 0.81 & 152.85&\\
LDA & 0.89 & 1.33&\\
QDA & 0.85 & 6.34&\\
SVM & 0.88 & -&0.84\\
Lasso & 0.89 & 1.29&0.87\\
Random Forest & 0.89 & 1.35&\\
\hline
\end{tabular}
\end{table}

## Chosen classifier

@@

We made the decision of model based on preliminary results, and though classification trees would represent an equally viable solution, we calculated those results only later on, and by that point had already calculated the results for the logistic regression model.

@@

The second problem is perhaps not as grave as the first. Especially in our challenge submission, our accuracy (0.84 for two classes and 0.68 for four) and perplexity (1.47) were not all that great. This was partially due to a coding error which made us choose a model that perhaps wasn't ideal, but there was still room for improvement even with our other models. We agreed before submitting the results that we are not aiming to win anything, and the submission itself is enough. We stand by this. We still could have done better both in numerical results and in predicting our own results, and we set to rectify this in our final report.

One of the other frustrations includes questions with our approach in relation to cross-validation. Perhaps the original choice of using loo-CV to assess the best model wasn't ideal and eventually in order to accommodate a lower amount of folds would have required us to possibly shift our entire approach from model accuracy and perplexity estimations to choice of ideal model and interpretation of said model as a result. Eventually we used a kind of messy half-measure where we decided our ideal model based on loo-CV and then for the model we used to predict the actual test data, we ran 10-fold cross validation to assess the ideal value of the tuning parameter. This was done to acknowledge the issues with the loo-CV even if, especially as we didn't enter into the contest with this model and for example the results in the table are still based on the loo-cv ideal model to compare it with the other models, and thus we didn't do much with the results we got.

Towards the 11th of December, we also found that many of our tree-based methods perform as well or better than logistic regression. If we had found this out earlier, we might have opted for a tree-based method as our final classifier. At that point we did not want to discard the work we had done so far, and opted to try to improve our logistic regression model instead. This decision proved to be the correct one, since the improved model outperformed our tree-based models.
